---
title: "Statistical description of iris data"
author: "Bryan Tchakote"
date: "12/12/2020"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    number_sections: yes
    toc: yes
---

```{r setup, include=FALSE}
library(moments)
knitr::opts_chunk$set(echo = TRUE)
```




# Description of the data set

```{r}
?iris
```

According to the help, **iris** data set gives the measurements in centimeters of the variables **sepal length and width and petal**
**length and width**, respectively, for 50 flowers from each of 3 species of iris. The species are **Iris setosa, versicolor, and**
**virginica**.

```{r}
dim(iris)
```
```{r}
colnames(iris)
```

**iris** is a data set of 150 rows (observations) and 5 columns (variables) which are:

- Sepal.Length
- Sepal.Width
- Petal.Length
- Petal.Width
- Species



# Univariate analysis

- Personnalized fonctions used for the univariate analysis

```{r Personnalized_fonctions_used_for_the_univariate_analysis}
numeric_summary = function(current_col){
  current_col_summary = c()
  
  assign_and_name = function(fun, name){
    x = fun
    names(x) = name
    current_col_summary = append(current_col_summary, x)
    return (current_col_summary)
  }
  
  # Measures of central location
  current_col_summary = assign_and_name(as.vector(summary(current_col)),
  names(summary(current_col)))
  
  
  # Measures of dispersion
  
  ## Range and Interquartile range
  current_col_summary = assign_and_name(current_col_summary[6] - current_col_summary[1],
  "Range")
  
  current_col_summary = assign_and_name(current_col_summary[5] - current_col_summary[2],
  "Interquartile range")
  
  ## Variance & Standard deviation
  current_col_summary = assign_and_name((var(current_col)*(length(current_col) - 1)) /
  length(current_col), "Variance")
  
  current_col_summary = assign_and_name(sd(current_col), "Standard deviation")
  
  
  # Skewness & Kurtosis
  current_col_summary = assign_and_name(skewness(current_col), "Skewness (Fisher)")
  current_col_summary = assign_and_name(kurtosis(current_col) - 3, "Kurtosis (Fisher)")
  
  current_col_summary = data.frame(current_col_summary)
  names(current_col_summary) = colnames(current_col)
  
  return (current_col_summary)
}

sk_and_kt_comments = function(sk, kt){
  if (sk < 0 ) {
    sk_c = "The distribution is negatively skewed, the mass of the
    distribution is concentrated on the right of the mean value.\n\n"
  } else if (sk > 0) {
    sk_c = "The distribution is positively skewed, the mass of the
    distribution is concentrated on the left of the mean value.\n\n"
  } else {
    sk_c = "The values are evenly distributed on both sides of the mean.\n\n"
  }
  
  if (kt < 0) {
    kt_c = "The distribution is platykurtic, according to the negative value of the
    kurtosis"
  } else if (kt > 0) {
    kt_c = "The distribution is leptokurtic, according to the positive value of
    the kurtosis"
  } else {
    kt_c = "The distribution is mesokurtic, the kurtosis value is zero"
  }
  
  return (c("Comment on the skewness\n", sk_c, "Comment on the kurtosis\n", kt_c))
}

graphic_summary = function(current_col){
  par(mfrow = c(1, 2))
  
  plot(table(current_col), xlab = colnames(current_col), ylab = "Occurences",
  col = "lightblue")
  
  boxplot(current_col, ylab = colnames(current_col), col = "lightblue",
  border = "darkgreen")
  
  par(mfrow = c(1, 1))
  
  plot(ecdf(current_col), main = colnames(current_col), col = "darkgreen")
}
```


## Sepal.Length

### Numeric summary

```{r}
S_L_summary = numeric_summary(iris[, 1])
S_L_summary
```

**Comments (about the skewness and the kurtosis especially):**

```{r}
sk_and_kt_comments(S_L_summary[[11, 1]], S_L_summary[[12, 1]])
```

### Graphic summary

```{r}
graphic_summary(iris[, 1])
```


## Sepal.Width

### Numeric summary

```{r}
S_W_summary = numeric_summary(iris[, 2])
S_W_summary
```

**Comments (about the skewness and the kurtosis especially):**

```{r}
sk_and_kt_comments(S_W_summary[[11, 1]], S_W_summary[[12, 1]])
```

### Graphic summary

```{r}
graphic_summary(iris[, 2])
```


## Petal.Length

### Numeric summary

```{r}
P_L_summary = numeric_summary(iris[, 3])
P_L_summary
```

**Comments (about the skewness and the kurtosis especially):**

```{r}
sk_and_kt_comments(P_L_summary[[11, 1]], P_L_summary[[12, 1]])
```

### Graphic summary

```{r}
graphic_summary(iris[, 3])
```


## Petal.Width

### Numeric summary

```{r}
P_W_summary = numeric_summary(iris[, 4])
P_W_summary
```

**Comments (about the skewness and the kurtosis especially):**

```{r}
sk_and_kt_comments(P_W_summary[[11, 1]], P_W_summary[[12, 1]])
```

### Graphic summary

```{r}
graphic_summary(iris[, 4])
```


## Species

```{r}
summary(iris$Species)
```



# Bivariate analysis


## Correlation between quantitatives variables

- Personnalized fonctions used for the bivariate analysis

```{r}
bivariate_analysis = function(col_X, col_Y){
  # Variables and respective marginal properties
  x_bar = mean(col_X)
  var_x = var(col_X) * ((length(col_X) - 1) / length(col_X))
  sd_x = sd(col_X)
  
  y_bar = mean(col_Y)
  sd_y = sd(col_Y)
  
  # Covariance
  cov_xy = cov(col_X, col_Y)
  
  # Correlation
  r_xy = cov_xy / (sd_x * sd_y)

  # Linear regression line (y = a + bx) - Coefficients a & b
  b = cov_xy / var_x
  a = y_bar - b*x_bar
  
  return (c(cov_xy, r_xy, a, b))
}

bivariate_graphs = function(X, Y, a, b){
  col_X = iris[, X]
  col_Y = iris[, Y]
  
  # Scatterplot
  plot(col_X, col_Y, col = "lightblue", pch = 20, xlab = names(iris)[X], ylab = names(iris)[Y],
  main = c("Scatterplot of", names(iris)[Y], "against", names(iris)[X]))
  
  # Linear regression line (actual line)
  abline(a, b, col = "darkgreen")
}
```


### Sepal.Length & Sepal.Width

Let:

- **Sepal.Length** be the *explanatory variable* (X)
- **Sepal.Width** be the *outcome variable* (Y)

```{r}
SL_SW_bivariate_analysis = bivariate_analysis(iris$Sepal.Length, iris$Sepal.Width)
```


#### Scatterplot

```{r}
# The scatterplot is going to be drawn lower, with the linear regression line
```

**Comment:** *A priori, the two variables seem to have a weak linear relationship. Let's verify it by calculating joint properties.*


#### Covariance & Correlation

- Covariance ($S_{xy}$)

```{r}
SL_SW_bivariate_analysis[1]
```

- Correlation coefficient ($r_{xy}$)

```{r}
SL_SW_bivariate_analysis[2]
```

**Comment:** *As expected, this correlation coefficient close to zero shows that the linear link between 'Sepal.Length' and 'Sepal.Width' is very weak*


#### Linear regression (y = a + bx)

```{r}
a = SL_SW_bivariate_analysis[3]
b = SL_SW_bivariate_analysis[4]

cat("Linear regression line equation:\ny =", a, "+", b, "x")

bivariate_graphs(1, 2, a, b)
```



### Petal.Length & Petal.Width

Let:

- **Petal.Length** be the *explanatory variable* (X)
- **Petal.Width** be the *outcome variable* (Y)

```{r}
PL_PW_bivariate_analysis = bivariate_analysis(iris$Petal.Length, iris$Sepal.Width)
```


#### Scatterplot

```{r}
# The scatterplot is going to be drawn lower, with the linear regression line
```

**Comment:** *A priori, the two variables seem to have a strong linear relationship. Let's verify it by calculating joint properties.*


#### Covariance & Correlation

- Covariance ($S_{xy}$)

```{r}
PL_PW_bivariate_analysis[1]
```

- Correlation coefficient ($r_{xy}$)

```{r}
PL_PW_bivariate_analysis[2]
```

**Comment:** *As expected, this correlation coefficient close to one shows that the linear link between 'Petal.Length' and 'Petal.Width' is very strong*


#### Linear regression (y = a + bx)

```{r}
a = PL_PW_bivariate_analysis[3]
b = PL_PW_bivariate_analysis[4]

cat("Linear regression line equation:\ny =", a, "+", b, "x")

bivariate_graphs(3, 4, a, b)
```




## Correlation between each quantitative variable and the only factor




### Sepal.Length & Species


#### Numeric summary

```{r}
S_L_summary_by_S = tapply(iris$Sepal.Length, iris$Species, numeric_summary)
S_L_summary_by_S = as.data.frame(S_L_summary_by_S, row.names = row.names(S_L_summary))

S_L_summary_by_S
```


#### Graphic summary

```{r}
tapply(iris$Sepal.Length, iris$Species, graphic_summary)
boxplot(iris$Sepal.Length ~ iris$Species, col = "lightblue", border = "darkgreen", xlab = "Species", ylab = names(iris)[1])

```